Fail Faster

### 2025-03-20 paper/paper.tex
wibbles

The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
inlined version takes (on average) \textbf{half} as much time to generate a single pair of
\texttt{int}s. 
Microbenchmarks of more realistic generators (see Section~\ref{section:eval}
\jwc{more precise ref}) show an even more dramatic performance boost for
inlining in this manner.

Because the inlined versions of the generator are identical to the un-inlined
versions except for mechanical, semantics-preserving transformations,
this performance difference is attributable solely to
the different machine code generated by the compiler.
Indeed, compilers of effectful and strict functional languages (including JIT
compilers, in the case of Scala 3)
use heuristics to determine if and when to perform this particular kind of
simplification~\footnote{As we discuss in Section~\ref{asdf} \jwc{fixme}, purity
means that Haskell is a slightly different story. GHC can and often does
transformations of this form.}.  Even in cases as simple as
Figure~\ref{fig:overhead-explanation-code}, the indirection of \texttt{return}
and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
even worse for recursive generators, as the heuristics are necessarily even more
conservative for optimizing recursive functions.

@@

While neither of these microbenchmarks exhibit this behavior, more complex
generators suffer further performance penalties due to \emph{closure
allocation}. In cases where the compiler cannot statically eliminate it,
running a monadic bind allocates a short-lived closure: we allocate a closure
for the continuation, and then immediately jump into it.
In strict functional languages like OCaml and Scala,
Each individual closure allocation is relatively cheap, but doing lots of allocation in a
generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
Similarly to \jwc{the previous one} this effect is magnified in recursive generators: each iteration
through the recursive loop re-allocates closures for \texttt{bind}s, so the amount of allocation
\emph{per generated value} scales linearly with the number of recursive generator calls.

\subsubsection{Overhead of Choice Combinators}
After the \texttt{return}, \texttt{bind}, and base-type generation combinators
(such as \texttt{gen\_int}), the most frequently used combinators are ``choice''
combinators like \texttt{weighted\_union} which randomly select between
sub-generators.
Like \texttt{return} and \texttt{bind}, these combinators incur a performance penalty
at run time. Aside from the previously-discussed issues that compilers
cannot see through the abstraction boundary to optimize these programs,
choice combinators come with their own particular abstraction overhead.

In practice, \texttt{weighted\_union} is almost always\footnote{
There are some generators where \texttt{weighted\_union} is passed a list which
was itself the result of a generator: the well-typed STLC term generator used in Section~\ref{section:eval} is an example of this.
} called with an explicitly constructed list, like
\texttt{weighted\_union [(w1,g1); (w2,g2); (w3,g3)]}. 
This is because the most common use case for \texttt{weighted\_union} is
to choose between one of the different constructors of an algebraic datatype,
the options for which are always known.
This list is allocated at the call site, and then \emph{immediately}
destructed by the function. In principle, since the elements of the list and its length $n$ are known statically,
a compiler could unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function to avoid allocating the list.
Unfortunately, almost~\footnote{Once again, GHC is a notable exception here, performing sophisticated list fusion optimizations \cn{}} no compilers perform this kind of optimization.
This allocation (or rather, its tendency to cause GC pauses) --- as well as the cost of running the code 
to traverse arbitrary lists compared to unrolled loops --- has a significant impact on 
performance.

>>>>

The performance impact of this inling is large (Figure~\ref{fig:overhead-explanation-perf}). In both languages, the
inlined version takes (on average) half as much time to generate a single pair of
\texttt{int}s. 
Microbenchmarks of more realistic generators (see Section~\ref{section:eval})
show an even more dramatic performance boost for inlining in this manner.

Because the inlined versions of the generator are identical to the un-inlined
versions except for mechanical, semantics-preserving transformations,
this performance difference is attributable solely to
the different machine code generated by the compiler.
Indeed, compilers of effectful and strict functional languages (including JIT
compilers, in the case of Scala 3)
use heuristics to determine if and when to perform this particular kind of
simplification~\footnote{As we discuss in Section~\ref{section:other-langs}, purity
means that Haskell is a slightly different story. GHC can and often does
transformations of this form.}.  Even in cases as simple as
Figure~\ref{fig:overhead-explanation-code}, the indirection of \texttt{return}
and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
even worse for recursive generators, as the heuristics are necessarily even more
conservative for optimizing recursive functions.

@@

While neither of these microbenchmarks exhibit this behavior, more complex
generators suffer further performance penalties due to \emph{closure
allocation}. In cases where the compiler cannot statically eliminate it,
running a monadic bind allocates a short-lived closure: we allocate a closure
for the continuation, and then immediately jump into it.
In strict functional languages like OCaml and Scala,
Each individual closure allocation is relatively cheap, but doing lots of allocation in a
generator is very costly because each allocation brings us closer to the next costly GC pause \cn{}.
This effect is also magnified in recursive generators: each iteration
through the recursive loop re-allocates closures for \texttt{bind}s, so the amount of allocation
\emph{per generated value} scales linearly with the number of recursive generator calls.

\subsubsection{Overhead of Choice Combinators}
Like \texttt{return} and \texttt{bind}, combinators like \texttt{weighted\_union} incur a performance penalty
at run time. Aside from the previously-discussed issues that compilers
cannot see through the abstraction boundary to optimize these programs,
choice combinators like \texttt{weighted\_union} come with their own particular abstraction overhead.

In practice, \texttt{weighted\_union} is almost always\footnote{
There are some generators where \texttt{weighted\_union} is passed a list which
was itself the result of a generator: the well-typed STLC term generator used in Section~\ref{section:eval} is an example of this.
} called with an explicitly constructed list, like
\texttt{weighted\_union [(w1,g1); (w2,g2); (w3,g3)]}. 
This is because the most common use case for \texttt{weighted\_union} is
to choose between one of the different constructors of an algebraic datatype,
the options for which are always known.
This list is allocated at the call site and then never needed after the call to \texttt{weighted\_union} returns\footnote{
This overhead is fixable in languages with stack alloctation, which OCaml will soon be \cn{}.
}.
Since the elements of the list and its length $n$ are known,
a compiler could in principle unroll the loops in \texttt{weighted\_union} to depth $n$ and specialize the function at each call site to avoid allocating the list.
Unfortunately, almost~\footnote{Once again, GHC is a notable exception here, performing sophisticated list fusion optimizations \cn{}} no compilers perform this kind of optimization.
This allocation (or rather, its tendency to cause GC pauses) --- as well as the cost of running the code 
to traverse arbitrary lists compared to unrolled loops --- has a significant impact on 
performance.
